---
title: "Who gets to work how?"
author: "Harald Kliems"
date: "12/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Prelude

I'll be using the CTPPr package to access data from the Census Transportation Planning Products. 
```{r}
library(tidyverse)
library(CTPPr)

```
First, I have to identify the relevant tables. I'm interested in means of transportation to work by race/ethnicity and by income. The `ctpp_tables` command provides an interactive way to identify the right tables. 
```{r identify-tables}
ctpp_tables()
```
However, I don't find this terribly user friendly and would rather use the CTPP [table browser](http://data5.ctpp.transportation.org/ctpp1216/Browse/browsetables.aspx).

It appears the two relevant tables are:

- `B102201` - Minority Status (3) by Means of Transportation (11) (Workers 16 years and over) 
- `B103203` - Household income in the past 12 months (2016 $) (26) by Means of transportation (11) (Workers 16 years and over in households)

For the latter, there is also "B103202 - Median Household Income in the last 12 months (2016 $) (1) by Means of Transportation (11) for Workers (Workers 16 years and over in households) - Large Geos Only" -- it's not clear what the difference is, and I will have to do some data exploration.

```{r download-data}
minority_commute <- download_ctpp(
  id = "B102201",
  dataset = "2016",
  geography = "City",
  state = "Wisconsin",
  output = "Name" #actual names
)

# fix columns names
minority_commute <- minority_commute %>% 
  rename("residence" = "RESIDENCE",
         "minority_status" = "Minority Status 3",
         "transpo_mode" = "Means of Transportation 11",
         "estimate" = "Estimate") %>% 
  mutate(minority_status = case_when(minority_status == "White alone, not Hispanic/Latino" ~ "non-Hispanic White",
                                     minority_status == "Other" ~ "Racial/ethnic minority",
                                     TRUE ~ minority_status))
```

Filter to only include Madison:
```{r}
minority_commute_msn <- minority_commute %>% 
  filter(residence == "Madison city, WI; Madison, WI Metro Area")
```

Minority status is dichotomized into non-Hispanic White and everyone else:
```{r}
unique(minority_commute_msn$minority_status)
```
Commute mode share is split into
```{r}
unique(minority_commute_msn$transpo_mode)
```
The table reports absolute numbers, which isn't super helpful:
```{r}
minority_commute_msn %>% 
  filter(transpo_mode == "Bicycle") %>% 
  ggplot(aes(x = minority_status, y = estimate)) +
  geom_col()
```

To get to percentages, we have to divide the estimates for each minority status by their respective totals. This requires separating the data out and then rejoining (at least I can't think of a more straightforward way):

```{r}
totals <- minority_commute_msn %>% 
  filter(transpo_mode == "Total, means of transportation") %>% 
  rename("total" = "estimate") %>% 
  select(minority_status, total)

minority_commute_msn <- minority_commute_msn %>% 
  left_join(totals, by = "minority_status") %>% 
  mutate(estimate_pct = estimate/total)
```
Quick faceted plot:
```{r}
library(hrbrthemes)
minority_commute_msn %>% 
  filter(transpo_mode != "Total, means of transportation") %>% 
  ggplot(aes(reorder(transpo_mode, estimate_pct), estimate_pct)) +
  geom_point() +
  geom_text(aes(label = round(estimate_pct, 2)*100), nudge_y = 0.1) +
  facet_wrap(~minority_status) +
  coord_flip() +
  theme_ipsum()
```
There are too many variables, including some that are pretty much at 0.

Let's combine:

```{r}
library(hrbrthemes)

minority_commute_msn %>%
  mutate(transpo_mode = fct_collapse(transpo_mode,
                                     "Drove alone" = "Car, truck, or van -- Drove alone",
                                     Bus = "Bus or trolley bus",
                                     Biked = "Bicycle",
                                     Carpooled = c("Car, truck, or van -- In a 2-person carpool",
                                                 "Car, truck, or van -- In a 3-or-more person carpool"),
                                     other = c("Railroad or ferryboat",
                                               "Streetcar, trolley car, subway, or elevated",
                                               "Taxicab, motorcycle or other method"))) %>% 
  filter(transpo_mode != "Total, means of transportation") %>%
  filter(transpo_mode != "other") %>% 
  group_by(minority_status, transpo_mode) %>% 
  summarize(pct = sum(estimate_pct)) %>% 
  ggplot(aes(x = reorder(transpo_mode, pct), y = pct, color = minority_status)) +
  geom_point(size = 3) +
  #geom_text(data = . %>% filter(minority_status == "Total persons"), aes(label = round(pct, 2)*100), nudge_x = 0.3) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_brewer(type = "qual",
                     palette = "Set3") +
  labs(x = element_blank(),
       y = "Commute mode share", 
       title = "Madison commute mode share by minority status",
       color = "Minority status")+
  #facet_wrap(~minority_status) +
  coord_flip() +
  theme_ft_rc() +
  theme(panel.grid.major.y = element_blank())
```

# By income
```{r}
income_commute <- download_ctpp(
  id = "B103203",
  dataset = "2016",
  geography = "City",
  state = "Wisconsin",
  output = "Name" #actual names
)
```

```{r}
income_commute <- income_commute %>% 
  rename("residence" = "RESIDENCE",
         "hh_income" = "Household Income in the Past 12 Months 2016 26",
         "transpo_mode" = "Means of Transportation 11",
         "estimate" = "Estimate") %>% 
  mutate(hh_income = ifelse(hh_income == "Total, household income", "Total", hh_income)) %>% 
  mutate(hh_income = factor(hh_income, unique(hh_income)))
```

```{r}
income_commute_msn <- income_commute %>% 
  filter(residence == "Madison city, WI; Madison, WI Metro Area")
```

```{r}
income_commute_msn %>% 
  ggplot(aes(hh_income, estimate, fill = transpo_mode)) +
  geom_col(position = "fill") +
  coord_flip()
```

```{r}
income_commute_msn %>% 
  mutate(hh_income = fct_collapse(hh_income,
                                  "< $15k" = c("Less than $5,000",
                                               "$ 5,000-$ 9,999",
                                               "$10,000-$12,499",
                                               "$12,500-$14,999"),
                                  "$15-25k" = c("$15,000-$17,499",
                                                "$17,500-$19,999",
                                                "$20,000-$22,499",
                                                "$22,500-$24,999"),
                                  "$25-50k" = c("$25,000-$27,499",
                                                "$27,500-$29,999",
                                                "$30,000-$32,499",
                                                "$32,500-$34,999",
                                                "$35,000-$37,499",
                                                "$37,500-$39,999",
                                                "$40,000-$42,499",
                                                "$42,500-$44,999",
                                                "$45,000-$47,499",
                                                "$47,500-$49,999"),
                                  "$50-75k" = c("$50,000-$54,999",
                                                "$55,000-$59,999",
                                                "$60,000-$74,999"),
                                  "$75-100k" = c("$75,000-$99,999"),
                                  "$100-125k" = c("$100,000-$124,999"),
                                  "$125-150k" = c("$125,000-$149,999"),
                                  ">$150k" = c("$150,000 or more"))) %>% 
  mutate(transpo_mode = fct_collapse(transpo_mode,
                                     "Drove alone" = "Car, truck, or van -- Drove alone",
                                     Bus = "Bus or trolley bus",
                                     Biked = "Bicycle",
                                     Carpooled = c("Car, truck, or van -- In a 2-person carpool",
                                                 "Car, truck, or van -- In a 3-or-more person carpool"),
                                     other = c("Railroad or ferryboat",
                                               "Streetcar, trolley car, subway, or elevated",
                                               "Taxicab, motorcycle or other method"))) %>%
  filter(!(transpo_mode %in% c("Total, means of transportation",
                               "other"))) %>% 
  group_by(hh_income, transpo_mode) %>% 
  summarize(estimate = sum(estimate)) %>% 
  mutate(transpo_mode = reorder(transpo_mode, estimate),
         transpo_mode = fct_relevel(transpo_mode, "Carpooled", after = 4)) %>% #move drove alone and carpooled next to each other
  ggplot(aes(hh_income, estimate, fill = transpo_mode)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(type = "qual",
                     palette = "Set3") +
  labs(x = element_blank(),
       y = "Commute mode share", 
       title = "Madison commute mode share by household income",
       fill = "Commute mode",
       caption = "Data: Census Transportation Planning Products 2012-16 
               Visualization: Harald Kliems")+
  #facet_wrap(~minority_status) +
  coord_flip() +
  theme_ft_rc() +
  theme(panel.grid.major.y = element_blank())
```

